\vspace{-0.5cm} XSGRB - The X-shooter GRB afterglow sample

### 2018-02-02 document/XSGRB_sample.tex
GCN citation and last comments, individual notes, Daniele contrib

In this paper, we present the results of a dedicated effort over the years 2009
-- 2017 to use the X-shooter spectrograph to secure spectroscopic observations
of afterglows and host galaxies of GRBs detected by {\it Swift}. We here make
all data resulting from the survey, publicly available in reduced form.

@@

\begin{enumerate}
	\item GRB triggered by BAT onboard the \swift~satellite 
	\item XRT started observing within 10 minutes after the GRB; an XRT position must be distributed within 12 hr.
	\item The target must be visible for at least 60 minutes, 30 degrees above the horizon, with the Sun below $-12$ degrees\footnote{Note that in the P84 proposal the criteria have
	been stated a bit differently, the visibility constraint being replaced by a
	declination + Sun angle constraint. These criteria are, however, those
	defining the sample.}.
	\item No bright, nearby stars within $ 1.8 + 0.4 \times exp[(20 - R)/2.05$] arcsec, where R is the USNO magnitude of the star)
	\item Galactic $A_V \lesssim 0.5$ mag.
\end{enumerate}

@@

Because the localization accuracy of BAT is 1 -- 4 arcminutes
\citep{Barthelmy2005}, a secure host association is in general very hard based
on BAT alone. We therefore additionally require an X-ray position from XRT to be
distributed to the GCN network \citep{Barthelmy2000} within 12 hours and to
account for observing constraints on \textit{Swift} that XRT began observations
within 10 minutes. The additional timing requirement of the XRT follow-up means
that all bursts in our sample have detected X-ray afterglows. Because the XRT
completeness is very high \citep{Burrows2007}, this cut should not alter the
parent sample significantly.

To ensure a minimum of observability, we require that the GRB is visible from
the telescope site at Cerro Paranal, Chile, for a least 1 hour after the trigger
with the sun below $-12$ degrees. This secures time for the spectroscopic
observations to be completed. Since the GRB population is isotropically
distributed on the sky as seen from Earth, and because the GRB properties do not
depend on position on the sky \citep{Meegan1992, Briggs1996, Ukwatta2016}, this
cut does not influence our ability to fairly sample the underlying GRB
population. The same arguments apply to the requirement that there are no nearby
bright stars. We additionally require that the Galactic extinction is below $A_V
\lesssim 0.5$ mag, based on the extinction maps by \citet{Schlegel1998,
	Schlafly2011}. Due to the isotropic distribution of GRBs on the sky, these
additional cuts should not influence the optical properties of the bursts
themselves, only our ability to successfully secure the observations that allows
us to investigate the spectroscopic properties of GRBs.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Follow-up procedure and practicalities}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\todo{Will be written by Daniele}

@@

Under rare circumstances, the use of the ESO rapid response mode (RRM) has been
possible. RRM is an automatic override of current observations at the telescope
that allows the shortest possible delays between the GRB trigger and initiation
of spectroscopic observations. This allows spectroscopic integration to begin on
the rapidly fading, optical transient within minutes of the burst. In case of a
promptly visible GRB, a robotic trigger was sent to the telescope if at the
moment of the GCN notice the GRB is observable and fulfills the following
criteria:

\begin{enumerate}
	\item GRB triggered by \textit{Swift}.
	\item X-shooter mounted at the telescope.
	\item GRB promptly visible.
	\item Error radius < 60\arcsec.
%	\item No limit on declination. 
%	\item No constraint on Moon phase or distance. 
	\item Elevation in the sky $> 22^\circ$ (both at trigger time and 15 min after).
	\item Sun elevation $< -12^\circ$ (both at trigger time and 15 min after).
	\item The X-ray position must be available less than 1 hr after the GRB.
	\item The \swift-circulated tags: KNOWN\_SOURCE, COSMIC\_RAY, DEF\_NOT\_GRB are set to false (PROB\_NOT\_GRB can be true)\footnote{These tags are distributed as part of the GCN notices. Explanations for all tags are \href{https://gcn.gsfc.nasa.gov/sock_pkt_def_doc.html}{available online}.}. 
\end{enumerate}

@@

There are nine GRBs in the sample that have been observed with the RRM mode.
One of the RRM triggers is outside the sample we use to address the statistical
properties of GRB afterglows and two of the RRM trigger are on short GRBs. The
fastest response between \swift~GRB trigger and the beginning of spectroscopic
integration is for GRB~160410A, for which the delay was only 8.4 minutes. In the
cases where the RRM trigger was rejected, it was in most cases due to a wrong
instrument being mounted at the focus.

Potentially, the proximity of the telescope (Paranal, Chile) to the South
Atlantic Anomaly (SSA) affects the rate of RRM triggers. BAT switches off when
the satellite goes through the SSA and that means we have less triggers of GRBs
that are immediately observable, compared to the rest of the \swift~orbit.

@@

The first GRB followed up was GRB~090313, observed on the 15th of March, 2009,
during the commissioning of X-shooter on the third Unit Telescope (UT3) of the
VLT . The bursts observed during the commissioning or the science verification
process (GRB~090313, GRB~090530, GRB~090809, GRB~090926) are not a part of the
sample we use to investigate the statistical properties of GRB afterglows, due
to the inhomogeneity of their selection. The first burst observed after science
verification and X-shooter was moved to UT2, was GRB~091018, which is the first
burst entering our statistically homogeneous sample. For all bursts that fulfill
our sample selection criteria, described in Sect. \ref{samplecrit},
spectroscopic follow-up have been attempted with X-shooter. Various conditions
can affect our ability to follow up a given burst, and a discussion of these
conditions and their consequences for the sample is included in Sect.
\ref{badbursts}.

@@

For a typical observation, each of the exposures in the nodding sequence has
been reduced as a single observation and then subsequently combined to form a
single image. We employ this strategy so that we can reject outliers in the
stack and weight by an averaged measure of the inverse variance of the
background. When weighting images, where the noise in each pixel is dominated by
Poisson noise, it is important to estimate the background variance in a large
enough region, so that any correlation between the signal and the weights are
removed. To this end, the weight map is generated by a running median window
over the variance map produced by the pipeline, where the trace has been masked
and the width of the window is chosen to be wide enough for the median variance to
be calculated on the basis of several hundred pixels. This weighting scheme
automatically also optimally combines images of different exposure times or
images where the background is varying, which is often the case when a burst has
been observed close to twilight

@@

In the NIR arm where the background is very bright and there are a high number
of broad sky-lines, an alternative approach to sky subtraction has been
employed. When there are no contaminating sources in the slit, the sky has been
put back on the images and the images are combined in pairs of two, subtracting
the two from each other while keeping the WCS static. Due to the nod-and-shuffle
offsets used between observations, this conserves the source flux while removing
the sky at the expense of a decrease in signal-to-noise by a factor of
$\sqrt{2}$. This amounts to the regular nodding reduction, only we can reject
outliers and weight by the averaged inverse variance map.

Reducing the images as single observations for all exposures, we additionally
get a spectrum of the sky which we can use to recalibrate the wavelength
solution.

@@

Since X-shooter is installed at the VLT Cassegrain focus, it is prone to
flexures. The flexures modify the projection of the slit on the detector with
respect to the one obtained in daytime calibration. This requires a modification
of the wavelength solution in order to correctly process the night-time data.
Part of this correction is performed by the pipeline using the frames taken
during the X-shooter Active Flexure Compensation procedure\footnote{X-shooter
	User Manual available at \href{https://www.eso.org/sci/facilities/paranal/instruments/xshooter/doc.html}{ESO website}}. The
remaining offset is corrected by cross-correlating the observed sky spectrum
with a synthetic sky spectrum \citep{Noll2012, Jones2013} after the continuum,
estimated as the mode of all flux values, has been subtracted. To get the
correct seeing point-spread-function (PSF) with which to convolve the synthetic sky, an initial
refinement of the wavelength solution has been obtained by cross-correlating the
observed sky with an unconvolved synthetic sky. This preliminary wavelength
calibration is applied to the observed sky. The synthetic spectrum is then
convolved with an increasing seeing PSF and the width that minimizes $\chi^2$
with the updated observed sky is chosen to be the effective sky-PSF. Using the
synthetic sky with the matched resolution, a final wavelength calibration can
then be calculated by cross-correlating the observed sky with the correctly
broadened sky spectrum as a function of a velocity offset. Both a
multiplicative and an additive offset to the wavelength calibration has been
tested, but in terms of $\chi^2$, the model with only a multiplicative offset is
preferred. The resulting offsets, which were smaller than 0.01 nm in the UVB and
VIS data and smaller than 0.05 nm in the NIR spectra, but changing over short
periods of time were applied to the corresponding spectra\footnote{The
	wavelength shifts have been extensively studied by
	ESO \href{https://www.eso.org/sci/facilities/paranal/instruments/xshooter/doc/XS_wlc_shift_150615.pdf}{in this document}.}. Using the convolved synthetic sky, the pixels containing the brightest sky lines have been added to the bad pixel map.

@@

To guide the estimated position of the trace on the slit as a function of
wavelength, we have used the analytic prescription for the trace position
described in \citet{Filippenko1982}, where the header keywords of the
observations have been queried for the ambient conditions which controls the
degree to which the trace changes position in the spatial direction.

Based on the signal-to-noise of the afterglow continuum, the 2D-image has been
binned down in the spectral direction to a number of elements that allows for an
accurate tracing of the PSF, typically 200 bins for moderate signal-to-noise.
For each of the bins, using the analytically guided guess position, the spectral
PSF has been fit using the unweighted chi-squared minimization algorithm
implemented in \texttt{scipy.optimize.curve\_fit} \citep{scipy}. Since we know
that the trace varies slowly as a function of wavelength, we have then fitted a
low-order polynomial to the fit parameters as a function of wavelength, which
allows us to evaluate the spectral PSF at all wavelengths and in this way
accurately model the entire spectral PSF.

@@

In cases where multiple traces are visible in the slit, additional components
for the profile are used in the optimal extraction. The additional components do
not share the PSF parameters and in cases where the additional component is an
extended object, the fits have been inspected to ensure that the additional
component does now skew the fit towards a different PSF. The additional
components are not used for the weights in the extractions.

The spectra are corrected for Galactic extinction using the $E(B-V)$ value from
the dust maps of \citet{Schlegel1998} with the update in
\citet{Schlafly2011}\footnote{Queried from
	http://irsa.ipac.caltech.edu/applications/DUST/index.html using \texttt{astroquery}
	\citep{astroquery}.}, and the extinction curve by \cite{Cardelli1989} with a
total to selective extinction $R_V = 3.1$. The wavelengths of the extracted
1D-spectra are moved to vacuum, corrected for barycentric motion, and the
wavelength recalibration described in Sect. \ref{wavecal} is applied. Pixels
with pixel-to-pixel variation larger than $50 \sigma$ are additionally added to
the bad pixel map.

@@

To this end we use observations of telluric standard stars that are taken with
identical instrument settings as our afterglow spectra, usually just after the
science data, as part of the ESO X-shooter calibration plan. These spectra have
been reduced together with the afterglow spectra, using identical pipeline
settings with the same version of the pipeline. First we fit a Gaussian function
in the spatial direction of the trace of the standard star at 792 nm (i.e. in
the VIS arm). After this, we fit a series of  20 telluric absorption lines in
the telluric standard star spectra with Gaussians, taking care to select
transitions that are not almost-resolved multiples, should be intrinsically
unresolved, and are in areas with well defined continuum flux. We pick 34
telluric standard stars spanning a range of DIMM seeing values, with the
majority between $0.5-1.5 $\arcsec. The resulting distribution of spectral FWHM
(km/s) as a function of spatial FWHM at 792 nm is fairly well described by a
linear relation $a + b*x$, with $x$ the spatial FWHM in pixels (with 0.15 arc
sec per pixel),  $a= 21.4 \pm 1.3$ km/s, $b=1.4 \pm 0.2$. We use this linear
relation as a way to estimate the spectral resolution for medium to low signal
to noise afterglow spectra in the VIS arm. To extend this to the UVB and NIR
arm, we measured a series of lines in the NIR arm spectra of a subset of 19 sources
used for the VIS arm above, and find that the resulting distribution is
consistent with a simple scaling of the VIS arm relation by the ratio of
resolutions of the NIR and VIS arm for unresolved, slit filling, sources as
given on the ESO instrument website. The UVB arm contains no suitable absorption
lines to use, and we therefore use a scaled value as in the NIR arm. This simple
analysis gives a sufficiently accurate estimate for the analysis of the low
signal-to-noise science spectra.

@@

All the spectra are made available as a single
\href{http://www.dark-cosmology.dk/~jselsing/XSGRB}{ZIP file}, through
\url{http://grbspec.iaa.es}, and additionally through the ESO archive in the
form of \href{http://archive.eso.org/wdb/wdb/adp/phase3_main/form}{phase 3
	material}. This release includes both prompt afterglow observations as well as
late time observations of the associated hosts, and represents \textit{all}
afterglow spectra of GRBs carried out by the X-shooter spectrograph since the
commissioning of the instrument, \startdate, and until the end of the last
period of the program 098.A-0055, \termdate~and thus constitutes eight years of GRB
afterglow observations with X-shooter. An overview of all the spectra and their
observational setups are given in Table \ref{tab:sample_overview}. For each
burst, each individual observation is provided in a separate reduction, and in
cases where observations have been repeated for an increased signal-to-noise or
to follow the temporal evolution, a combined spectrum is also provided. No
attempt has been made to join the spectroscopic arms, so for each observation,
three spectra are provided in separate files.

@@

Each file contains 7 columns with the following contents and descriptions:
\begin{itemize}
	\item WAVE - Observed wavelength in vacuum, corrected for barycentric motion and drifts in the wavelength solution ($\mathrm{\AA}$).
	\item FLUX - Observed flux density ($\mathrm{erg} \mathrm{s}^{-1} \mathrm{cm}^{-2} \mathrm{\AA}^{-1}$).
	\item ERR - Associated flux density error ($\mathrm{erg} \mathrm{s}^{-1} \mathrm{cm}^{-2} \mathrm{\AA}^{-1}$).
	\item QUAL - Bad pixel map, where a value different from zero indicates a bad pixel.
	\item CONTINUUM - Continuum estimate based on Sect. \ref{continuum} ($\mathrm{erg} \mathrm{s}^{-1} \mathrm{cm}^{-2} \mathrm{\AA}^{-1}$).
	\item CONTINUUM\_ERR - Relative error on the continuum estimate.
	\item TELL\_CORR - Inverse transmission spectrum. Multiply FLUX and ERR column with this column to correct for telluric absorption. 
\end{itemize}

@@

In this section, we describe the efficiency of the follow-up effort and the
characteristics of the observed bursts. We also assess the degree to which the
obtained sample is representative for the full \textit{Swift} sample. An
important note is that here we include \textit{all} GRBs after \startdate, that
have been observed with X-shooter, while only a subset of these constitutes our
\textit{statistical} sample. The statistical sample is based on the selection
criteria described in Sect. \ref{samplecrit}. Some bursts not fulfilling the
sample criteria have been followed up due to interesting characteristics, e.g.,
curios properties of their light curves, their brightness, etc. These bursts
are not discussed as part of the investigation of the statistical properties of
the GRB population. A prime example of a spectacular burst outside the
statistical sample is the bright \textit{INTEGRAL} burst GRB~161023A, that
contains at least 11 intervening absorption systems (See \ref{161023}).

@@

Redshift determination of bursts for which the host is too faint for a
spectroscopic redshift measurement relies on the detection of absorption lines
imprinted on the GRB afterglow continuum. Because the afterglow rapidly fades
(typically as $\sim t^{-1}$) a rapid follow-up is essential. In Fig.
\ref{fig:timing} we plot the delay from the BAT trigger to the start of the
spectroscopic observation. The shortest delays are observed in RRM-mode. The
fastest follow-up between BAT trigger and start of spectroscopic observations
for any observation is for the short, $z = 1.717$, GRB~160410A for which
spectroscopic integration was initiated only 8.4 minutes after the BAT trigger.
To illustrate the importance of the follow-up delay for the redshift
completeness, we plot the redshift completeness as a function of delay time in
Fig. \ref{fig:timing} for all the bursts we have followed up, including the ones
outside the statistical sample. As can be seen from the figure, the fraction of
GRBs with a redshift determination decreases with follow-up delay. The redshift
completeness for bursts that we have followed up is 92 per cent. This degree of
completeness in the followed bursts, illustrates the efficiency of VLT/X-shooter
in redshift determination. Not shown in the figure are an additional 12 bursts
that have redshift determinations based on late-time host observations with
delay times longer than $\sim$10 days.

@@

\bibliographystyle{mnras}
\bibliography{XSGRB_sample}

@@

Observed 3 days after the \textit{Swift} trigger. No afterglow or host detected
in 10.8 ks. GRB likely located at high
redshift\footnote{\url{http://gcn.gsfc.nasa.gov/gcn3/10399.gcn3}}. The spectrum
has not otherwise been published previously.

>>>>

In this paper, we present the results of a dedicated effort over the years 2009
-- 2017 to use the X-shooter spectrograph to secure spectroscopic observations
of afterglows and host galaxies of GRBs detected by {\it Swift}. We here make
all data resulting from the survey, publicly available in reduced form (see Sect
\ref{products}).

@@

\begin{enumerate}

	\item GRB triggered by BAT onboard the \swift~satellite

	\item XRT started observing within 10 minutes after the GRB; an XRT position
	must be distributed within 12 hr.

	\item The target must be visible for at least 60 minutes, 30 degrees above the
	horizon, with the Sun below $-12$ degrees\footnote{Note that in the P84
		proposal the criteria have been stated differently, the visibility constraint
		being replaced by a declination + Sun angle constraint. The criteria listed
		above are, however, those defining the sample.}.

	\item No bright, nearby stars within $ 1.8 + 0.4 \times \exp[(20 - R)/2.05$]
	arcsec, where $R$ is the USNO magnitude of the star)

	\item Galactic $A_V \lesssim 0.5$ mag according to the maps of \citet{Schlegel1998}.

\end{enumerate}

@@

Because the localization accuracy of BAT is 1 -- 4 arcminutes
\citep{Barthelmy2005}, an afterglow identification based on BAT alone would be
harder, and  host association impossible. We therefore additionally require an
X-ray position from XRT to be distributed to the GCN network
\citep{Barthelmy2000} within 12 hours and to account for observing constraints
on \textit{Swift} that XRT began observations within 10 minutes. The additional
timing requirement of the XRT follow-up means that all bursts in our sample have
detected X-ray afterglows. Because the XRT completeness is very high for
promptly-repointed GRBs \citep{Burrows2007}, this cut should not alter the
parent sample significantly.

To ensure a minimum of observability, we require that the GRB is visible from
the telescope site at Cerro Paranal, Chile, for a least 1 hour after the trigger
with the sun below $-12$ degrees. This secures time for the spectroscopic
observations to be completed. Since the GRB population is isotropically
distributed on the sky as seen from Earth, and because the GRB properties do not
depend on position on the sky \citep{Meegan1992, Briggs1996, Ukwatta2016}, this
cut does not influence our ability to fairly sample the underlying GRB
population. The same arguments apply to the requirement that there are no nearby
foreground bright stars. We additionally require that the Galactic extinction is
below $A_V \lesssim 0.5$ mag, based on the extinction maps by
\citet{Schlegel1998}\footnote{We use the updated values of \citet{Schlafly2011}
	to correct for foreground extinction, but the sample criterion is based on the
	old \citet{Schlegel1998} values for consistency with the first semesters of our
	program.}. Choosing low-extinction sightlines also reduce the problem of field
crowding and the contamination from Galactic outbursts posing as GRB impostors.
These additional cuts should not influence the optical properties of the bursts
themselves, only our ability to successfully secure the observations that allows
us to investigate the spectroscopic properties of GRBs.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Follow-up procedure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Our collaboration has set up a procedure to promptly react to alerts from
\swift~and other satellites (such as \textit{Fermi}). We process automatically
GCN notices in order to flag those events belonging to our statistical sample.
Two people are permanently on alert to manually supervise each event, for
example to recognize events of special interest beyond their inclusion in the
sample. Members of our collaboration have access to numerous facilities spread
throughout the world, among which the Nordic Optical Telescope (NOT), the Gran
Telescopio Canarias (GTC), the Telescopio Nazionale Galileo (TNG), the Gamma-Ray
Burst Optical Near-infrared Detector (GROND), the Xinlong observatory. Thanks to
this network, we can often conduct searches for optical/NIR afterglows ahead of
the time of the X-shooter spectroscopy, thus helping to plan the spectroscopic
observations. If this is not possible, we use the built-in acquisition camera of
X-shooter to search for counterparts, and adapt the observing strategy on the
fly. In many cases we could interact directly with the staff at the telescope to
aid the observation. Raw data are usually available within minutes in the ESO
archive LINK!!! and are promptly reduced by members of our collaboration, often
using archival calibration data which are readily available. This allows us in
most cases to report the preliminary results (redshift, identification of the
most prominent emission and absorption feature, etc.) within a few hours after
the beginning of the observations.

@@

Under rare circumstances, the use of the ESO rapid response mode (RRM;
\citealt{Vreeswijk2010}) has been possible. RRM is a system which automatically
overrides current observations at the telescope. This allows the shortest
possible delay between the GRB trigger and initiation of observations. This
allows spectroscopic integration to begin on the rapidly fading, optical
transient within minutes of the burst. In case of a promptly visible GRB, a
robotic trigger was sent to the telescope if at the moment of the GCN notice the
GRB is observable and fulfills the following criteria:

\begin{enumerate}
	\item GRB triggered by \textit{Swift}.
	\item X-shooter is the currently operating instrument at the telescope.
	\item GRB promptly visible from Paranal.
	\item Error radius < 60\arcsec.
%	\item No limit on declination. 
%	\item No constraint on Moon phase or distance. 
	\item Elevation in the sky $> 22^\circ$ (both at trigger time and 15 min after).
	\item Sun elevation $< -12^\circ$ (both at trigger time and 15 min after).
	\item The X-ray position must be available less than 1 hr after the GRB.
	\item The \swift-circulated tags: KNOWN\_SOURCE, COSMIC\_RAY, DEF\_NOT\_GRB are set to false (PROB\_NOT\_GRB can be true)\footnote{These tags are distributed as part of the GCN notices. Explanations for all tags are \href{https://gcn.gsfc.nasa.gov/sock_pkt_def_doc.html}{available online}.}. 
\end{enumerate}

The criteria to trigger RRM are looser than those to define the statistical
sample. This is because of the expected larger brightness of any GRB counterpart
minutes after the explosion, and because of the rarity of RRM triggers.

@@

There are nine GRBs that have been observed by us with the RRM mode. One of the
RRM triggers is outside the statistical sample  and two of the RRM trigger are
on short GRBs. The fastest response between \swift~GRB trigger and the beginning
of spectroscopic integration was for GRB~160410A, for which the delay was only
8.4 minutes. In many cases, unfortunately, we could not use RRM even for promptly visible events, because of the unavailability of X-shooter.

Potentially, the proximity of the telescope (Paranal, Chile) to the South
Atlantic Anomaly (SSA) affects the rate of RRM triggers \citep{Greiner2011}. BAT
switches off when the satellite goes through the SSA and that means we have
fewer GRB triggers immediately observable, compared to the rest of
the \swift~orbit.

@@

The first GRB followed up was GRB~090313 \citep{DeUgartePostigo2010}, observed
on the 15th of March, 2009, during the commissioning of X-shooter on the third
Unit Telescope (UT3) of the VLT . The bursts observed during the commissioning
or the science verification process (GRB~090313, GRB~090530, GRB~090809,
GRB~090926) are not a part of the sample we use to investigate the statistical
properties of GRB afterglows, due to the inhomogeneity of their selection. The
first burst observed after science verification and X-shooter was moved to UT2,
was GRB~091018, which is the first burst entering our statistically homogeneous
sample. For all bursts that fulfill our sample selection criteria, described in
Sect. \ref{samplecrit}, spectroscopic follow-up have been attempted with
X-shooter. Various conditions can affect our ability to follow up a given burst,
and a discussion of these conditions and their consequences for the sample is
included in Sect. \ref{badbursts}.

@@

For a typical observation, each of the exposures in the nodding sequence has
been reduced as a single observation and then subsequently combined to form a
single image. We employ this strategy so that we can reject outliers in the
stack and weight by an averaged measure of the inverse variance of the
background. When weighting images, where the noise in each pixel is dominated by
Poisson noise, it is important to estimate the background variance in a large
enough region, so that any correlation between the signal and the weights are
removed. To this end, the weight map is generated by a running median window
over the variance map produced by the pipeline, where the trace has been masked
and the width of the window is chosen to be wide enough for the median variance to
be calculated on the basis of several hundred pixels. This weighting scheme
automatically also optimally combines images of different exposure times or
images where the background is varying, which is often the case when a burst has
been observed close to twilight.

@@

In the NIR arm where the background is very bright and there are a high number
of broad sky-lines, an alternative approach to sky subtraction has been
employed. When there are no contaminating sources in the slit, the sky has been
put back on the images and the images are combined in pairs of two before
extraction, subtracting the two from each other while keeping the WCS static.
Due to the nod-and-shuffle offsets used between observations, this conserves the
source flux while removing the sky at the expense of a decrease in
signal-to-noise by a factor of $\sqrt{2}$. This amounts to the regular nodding
reduction, only we can reject outliers and weight by the averaged inverse
variance map.

Reducing the images as single observations for all exposures, we additionally
get a spectrum of the sky which we can use to recalibrate the wavelength
solution in the post-processing steps.

@@

Since X-shooter is installed at the VLT Cassegrain focus, it is prone to
flexures. The flexures modify the projection of the slit on the detector with
respect to the one obtained in daytime calibration. This requires a modification
of the wavelength solution in order to correctly process the night-time data.
Part of this correction is performed by the pipeline using the frames taken
during the X-shooter Active Flexure Compensation procedure\footnote{X-shooter
	User Manual available at \href{https://www.eso.org/sci/facilities/paranal/instruments/xshooter/doc.html}{ESO website}.}. The
remaining offset is corrected by cross-correlating the observed sky spectrum
with a synthetic sky spectrum \citep{Noll2012, Jones2013} after the continuum,
estimated as the mode of all flux values, has been subtracted. To get the
correct seeing point-spread-function (PSF) with which to convolve the synthetic sky, an initial
refinement of the wavelength solution has been obtained by cross-correlating the
observed sky with an unconvolved synthetic sky. This preliminary wavelength
calibration is applied to the observed sky. The synthetic spectrum is then
convolved with an increasing seeing PSF and the width that minimizes $\chi^2$
with the updated observed sky is chosen to be the effective sky-PSF. Using the
synthetic sky with the matched resolution, a final wavelength calibration can
then be calculated by cross-correlating the observed sky with the correctly
broadened sky spectrum as a function of a velocity offset. Both a
multiplicative and an additive offset to the wavelength calibration has been
tested, but in terms of $\chi^2$, the model with only a multiplicative offset is
preferred. The resulting offsets, which were smaller than 0.01 nm in the UVB and
VIS data and smaller than 0.05 nm in the NIR spectra, but changing over short
periods of time were applied to the corresponding spectra\footnote{The
	wavelength shifts have been extensively studied by ESO
	\href{https://www.eso.org/sci/facilities/paranal/instruments/xshooter/doc/XS_wlc_shift_150615.pdf}{in this document}.}. Using the convolved synthetic sky, the pixels containing the brightest sky lines have been flagged as such in the bad pixel map.

@@

To guide the estimated position of the trace on the slit as a function of
wavelength, we have used the analytic prescription for the trace position
described in \citet{Filippenko1982}, where the header keywords of the
observations have been queried for the ambient conditions which controls the
degree to which the trace changes position in the spatial direction. This
analytic approach is only valid for a plane-parallel atmosphere, but because the
final position is refined in the fit, it is adequate for our purposes.

Based on the signal-to-noise of the afterglow continuum, the 2D-image has been
binned in the spectral direction to a number of elements that allows for an
accurate tracing of the PSF, typically 200 bins for moderate signal-to-noise.
For each of the bins, using the analytically guided guess position, the spectral
PSF has been fit using the unweighted chi-squared minimization algorithm
implemented in \texttt{scipy.optimize.curve\_fit} \citep{scipy}. Since we know
that the trace varies slowly as a function of wavelength, we have then fitted a
low-order polynomial to the fit parameters as a function of wavelength, which
allows us to evaluate the spectral PSF at all wavelengths and in this way
accurately model the entire spectral PSF.

@@

In cases where multiple traces are visible in the slit, additional components
for the profile are used in the optimal extraction. The additional components do
not share the PSF parameters and in cases where the additional component is an
extended object, the fits have been inspected to ensure that the additional
component does not skew the fit towards a different PSF. The additional
components are not used for the weights in the extractions.

The spectra are corrected for Galactic extinction using the $E(B-V)$ value from
the dust maps of \citet{Schlegel1998} with the update in
\citet{Schlafly2011}\footnote{Queried from
	http://irsa.ipac.caltech.edu/applications/DUST/index.html using
	\texttt{astroquery} \citep{astroquery}.}, and the extinction curve by
\cite{Cardelli1989} with a total to selective extinction $R_V = 3.1$. The
wavelengths of the extracted 1D-spectra are wavelength recalibrated (described
in Sect. \ref{wavecal}), moved to vacuum, and corrected for barycentric motion.
Pixels with pixel-to-pixel variation larger than $50 \sigma$ are additionally
added to the bad pixel map.

@@

To this end we use observations of telluric standard stars that are taken with
identical instrument settings as our afterglow spectra, usually just after the
science data, as part of the ESO X-shooter calibration plan. These spectra have
been reduced together with the afterglow spectra, using identical pipeline
settings with the same version of the pipeline. First we fit a Gaussian function
in the spatial direction of the trace of the standard star at 792 nm (i.e. in
the VIS arm). After this, we fit a series of  20 telluric absorption lines in
the telluric standard star spectra with Gaussians, taking care to select
transitions that are not almost-resolved multiples, should be intrinsically
unresolved, and are in areas with well defined continuum flux. We pick 34
telluric standard stars spanning a range of DIMM seeing values, with the
majority between $0.5-1.5 $\arcsec. The resulting distribution of spectral FWHM
(km/s) as a function of spatial FWHM at 792 nm is fairly well described by a
linear relation $a + b*x$, with $x$ the spatial FWHM in pixels (with 0.15 arc
sec per pixel),  $a= 21.4 \pm 1.3$ km/s, $b=1.4 \pm 0.2$ (see Fig.
\ref{fig:res}). We use this linear relation as a way to estimate the spectral
resolution for medium to low signal to noise afterglow spectra in the VIS arm.
To extend this to the UVB and NIR arm, we measured a series of lines in the NIR
arm spectra of a subset of 19 sources used for the VIS arm above, and find that
the resulting distribution is consistent with a simple scaling of the VIS arm
relation by the ratio of resolutions of the NIR and VIS arm for unresolved, slit
filling, sources as given on the ESO instrument website. The UVB arm contains no
suitable absorption lines to use, and we therefore use a scaled value as in the
NIR arm. This simple analysis gives a sufficiently accurate estimate for the
analysis of the low signal-to-noise science spectra.

@@

All the spectra are made available as a single
\href{http://www.dark-cosmology.dk/~jselsing/XSGRB}{ZIP file}, through
\url{http://grbspec.iaa.es}, and additionally through the ESO archive in the
form of \href{http://archive.eso.org/wdb/wdb/adp/phase3_main/form}{phase 3
	material}. This release includes both prompt afterglow observations as well as
late time observations of the associated hosts, and represents \textit{all}
afterglow spectra of GRBs carried out by the X-shooter spectrograph since the
commissioning of the instrument, \startdate, and until the end of the last
period of the program 098.A-0055, \termdate~and thus constitutes eight years of GRB
afterglow observations with X-shooter. An overview of all the spectra and their
observational setups is given in Table \ref{tab:sample_overview}. For each
burst, each individual observation is provided in a separate reduction, and in
cases where observations have been repeated for an increased signal-to-noise or
to follow the temporal evolution, a combined spectrum is also provided. No
attempt has been made to join the spectroscopic arms, so for each observation,
three spectra are provided in separate files.

@@

Each file contains 7 columns with the following contents and descriptions:
\begin{itemize}

	\item WAVE - Observed wavelength in vacuum, corrected for barycentric motion
	and drifts in the wavelength solution ($\mathrm{\AA}$).

	\item FLUX - Observed flux density ($\mathrm{erg} \mathrm{s}^{-1}
	\mathrm{cm}^{-2} \mathrm{\AA}^{-1}$).

	\item ERR - Associated flux density error ($\mathrm{erg} \mathrm{s}^{-1}
	\mathrm{cm}^{-2} \mathrm{\AA}^{-1}$).

	\item QUAL - Bad pixel map, where a value different from zero indicates a bad
	pixel.

	\item CONTINUUM - Continuum estimate based on Sect. \ref{continuum}
	($\mathrm{erg} \mathrm{s}^{-1} \mathrm{cm}^{-2} \mathrm{\AA}^{-1}$).

	\item CONTINUUM\_ERR - Relative error on the continuum estimate.

	\item TELL\_CORR - Inverse transmission spectrum. Multiply FLUX and ERR column
	with this column to correct for telluric absorption.

\end{itemize}

@@

In this section, we describe the efficiency of the follow-up effort and the
characteristics of the observed bursts. We also assess the degree to which the
obtained sample is representative for the full \textit{Swift} sample. An
important note is that here we include \textit{all} GRBs after \startdate, that
have been observed with X-shooter, while only a subset of these constitutes our
\textit{statistical} sample. The statistical sample is based on the selection
criteria described in Sect. \ref{samplecrit}. Some bursts not fulfilling the
sample criteria have been followed up due to interesting characteristics, e.g.,
curios properties of their light curves, their brightness, etc. These bursts
are not discussed as part of the investigation of the statistical properties of
the GRB population. A prime example of a spectacular burst outside the
statistical sample is the bright \textit{INTEGRAL} burst GRB~161023A, that
contains at least 15 intervening absorption systems (See \ref{161023}).

@@

Redshift determination of bursts for which the host is too faint for a
spectroscopic redshift measurement relies on the detection of absorption lines
imprinted on the GRB afterglow continuum. Because the optical afterglow rapidly
fades (typically as $\sim t^{-1}$) a rapid follow-up is essential. In Fig.
\ref{fig:timing} we plot the delay from the BAT trigger to the start of the
spectroscopic observation. The shortest delays are observed in RRM-mode. The
fastest follow-up between BAT trigger and start of spectroscopic observations
for any observation is for the short, $z = 1.717$, GRB~160410A for which
spectroscopic integration was initiated only 8.4 minutes after the BAT trigger.
To illustrate the importance of the follow-up delay for the redshift
completeness, we plot the redshift completeness as a function of delay time in
Fig. \ref{fig:timing} for all the bursts we have followed up, including the ones
outside the statistical sample. As can be seen from the figure, the fraction of
GRBs with a redshift determination decreases with follow-up delay. The redshift
completeness for bursts that we have followed up is 92 per cent. This degree of
completeness in the followed bursts, illustrates the efficiency of VLT/X-shooter
in redshift determination. Not shown in the figure are an additional 12 bursts
that have redshift determinations based on late-time host observations with
delay times longer than $\sim$10 days.

@@

\bibliographystyle{mnras}
\bibliography{XSGRB_sample,GCN}

@@

Observed 3 days after the \textit{Swift} trigger. No afterglow or host detected
in 10.8 ks. GRB likely located at high
redshift \citep[GCN GCN10399;][]{GCN10399}. The spectrum
has not otherwise been published previously.
